---
title: "Reproducible Research: Peer Assessment 1"
author: "Vitalii Kleshchevnikov"
output: 
  html_document:
    keep_md: true
---

Running this Rmd requires tidyr and ggplot packages

## Loading and preprocessing the data

Reading data - and quick look. 
``` {r loading} 
# setwd("./RepData_PeerAssessment1")
activity.dat = read.csv("activity.csv", colClasses = c("numeric", "Date","integer"),stringsAsFactors = F)
head(activity.dat)
str(activity.dat)
```
"spreading" data frame - so that each row contains all observations for each day,
intervals in columns
```{r transforming}
activity.dat.1row.per.date = tidyr::spread(activity.dat, interval, steps)
activity.dat.1row.per.interval = tidyr::spread(activity.dat, date, steps)
# cheking how data was spread - code below produces lond output - disabled
# str(activity.dat.1row.per.interval)
# str(activity.dat.1row.per.date)
```


## What is mean total number of steps taken per day?

Removing days which doesn't have any measurements and
calculating total number of steps per day (NA omitted) - plotting histogram
```{r steps_per_day}
activity.dat.1row.per.date.na.rm = activity.dat.1row.per.date[complete.cases(activity.dat.1row.per.date),]
total_steps.per.day = rowSums(activity.dat.1row.per.date.na.rm[,2:289])

library(ggplot2)
qplot(x=total_steps.per.day, geom = "histogram", bins = 15) +
      xlab("the total number of steps taken each day") +
      ggtitle("Histogram of the total number of steps taken each day") +
      theme(title = element_text(size = rel(1.2)))
```

```{r mean_steps.per.day}
mean_steps.per.day = format(round(mean(total_steps.per.day),0), scientific=F)
# round to the nearest integer value (steps cannot be fractional)
# format to avoid 1.07*10^4 representation
```
mean number of steps per day is `r mean_steps.per.day`

```{r median_steps.per.day}
median_steps.per.day = format(round(median(total_steps.per.day),0),scientific=F)
# round to the nearest integer value (steps cannot be fractional)
# format to avoid 1.07*10^4 representation
```
median number of steps per day is `r median_steps.per.day`


## What is the average daily activity pattern?
Calculating the mean number of steps per time interval (over all days) and
the interval containing maximum number of steps
```{r steps_per_interval}
mean_steps.per.interval = rowMeans(activity.dat.1row.per.interval[,2:62], na.rm = T)
max_step_interval = 
      activity.dat.1row.per.interval$interval[mean_steps.per.interval == max(mean_steps.per.interval)]
```

```{r plotting_steps_per_interval, fig.width=9}
qplot(x = activity.dat.1row.per.interval$interval, y = mean_steps.per.interval, geom="path") + 
      xlab("time interval") +
      ylab("mean number of steps")+
      ggtitle("the average (mean) number of steps taken in different intervals during the day") + scale_x_continuous(breaks = seq(0,2400,100))+
      theme(title = element_text(size = rel(1)))
```

The interval containing maximum number of steps (on average(mean) across all days) - `r max_step_interval`


## Imputing missing values

```{r N_missing_observations}
missing_rows = sum(!complete.cases(activity.dat))
```
the number of intervals missing observations (the number of steps) is `r missing_rows`

Imputing missing values by setting each missing observation (steps) to be equal 
to the mean number of steps taken in that interval
```{r imputing missing values}
activity.dat.1row.per.interval.imputed = activity.dat.1row.per.interval
activity.dat.1row.per.interval.imputed[is.na(activity.dat.1row.per.interval)] = mean_steps.per.interval
```
Calculating total number of steps per day (NA imputed) - plotting histogram
```{r histogram_steps_day}
total_steps.per.day.imputed = 
      colSums(activity.dat.1row.per.interval.imputed[,2:62], na.rm = F)

library(ggplot2)
qplot(x=total_steps.per.day.imputed, geom = "histogram", bins = 15) + 
      xlab("the total number of steps taken each day") +
      ggtitle("Histogram of the total number of steps taken each day \n (after missing values were imputed)") +
      theme(title = element_text(size = rel(1.2)))
```

```{r mean_imputed}
mean_steps.per.day.imputed = format(round(mean(total_steps.per.day.imputed),0),scientific=F)
# round to the nearest integer value (steps cannot be fractional)
# format to avoid 1.07*10^4 representation
```
mean number of steps per day doesn't change after imputing and equals to `r mean_steps.per.day.imputed`.

```{r median_imputed}
median_steps.per.day.imputed = format(round(median(total_steps.per.day.imputed),0),scientific=F)
# round to the nearest integer value (steps cannot be fractional)
# format to avoid 1.07*10^4 representation
```
Median number of steps per day after imputing is `r median_steps.per.day`. 
Imputing missing values doesn't substantially change median either.

## Are there differences in activity patterns between weekdays and weekends?

Code below transfors the data.frame with imputed values into the form of original 
dataset, adds the fourth variable weekday/weekend, separates data.frame by 
weekday/weekend, transforms data.frame by spreading, recalculates the mean 
number of steps in each time interval over weekdays or weekends. 
Plots the result: during the day on x-axis, average(mean) number of steps on the 
y-axis)

```{r weekdays_vs_weekends, fig.width= 9, fig.height=6}
activity.dat.weekdays = tidyr::gather(data = activity.dat.1row.per.interval.imputed,
                                      key = date, value = steps, -interval)
activity.dat.weekdays[,4] = weekdays(as.Date(activity.dat.weekdays$date))
colnames(activity.dat.weekdays)[4] = "weekday"
# Changing specific days to weekday/weekend
activity.dat.weekdays$weekday = gsub("Friday", "weekday", activity.dat.weekdays$weekday)
activity.dat.weekdays$weekday = gsub("Monday", "weekday", activity.dat.weekdays$weekday)
activity.dat.weekdays$weekday = gsub("Tuesday", "weekday", activity.dat.weekdays$weekday)
activity.dat.weekdays$weekday = gsub("Wednesday", "weekday", activity.dat.weekdays$weekday)
activity.dat.weekdays$weekday = gsub("Thursday", "weekday", activity.dat.weekdays$weekday)
activity.dat.weekdays$weekday = gsub("Saturday", "weekend", activity.dat.weekdays$weekday)
activity.dat.weekdays$weekday = gsub("Sunday", "weekend", activity.dat.weekdays$weekday)
# Transforming data for plotting
activity.dat.weekdays.day = dplyr::filter(activity.dat.weekdays, weekday == "weekday")
activity.dat.weekdays.day.s = tidyr::spread(activity.dat.weekdays.day, date, steps)
mean_steps.per.interval.weekday = data.frame(interval = activity.dat.weekdays.day.s$interval, 
                  mean_steps = rowMeans(activity.dat.weekdays.day.s[,3:ncol(activity.dat.weekdays.day.s)]), 
                  weekday = activity.dat.weekdays.day.s$weekday)

activity.dat.weekdays.end = dplyr::filter(activity.dat.weekdays, weekday == "weekend")
activity.dat.weekdays.end.s = tidyr::spread(activity.dat.weekdays.end, date, steps)
mean_steps.per.interval.weekend = data.frame(interval = activity.dat.weekdays.end.s$interval, 
                  mean_steps = rowMeans(activity.dat.weekdays.end.s[,3:ncol(activity.dat.weekdays.end.s)]), 
                  weekday = activity.dat.weekdays.end.s$weekday)

activity.weekdays = rbind(mean_steps.per.interval.weekday, mean_steps.per.interval.weekend)

# plotting
ggplot(activity.weekdays, aes(x=interval, y = mean_steps))+
      geom_path() + facet_grid(weekday~.)+
            xlab("time interval") +
      ylab("mean number of steps")+
      ggtitle("the average (mean) number of steps taken in different intervals \n during the day, weekdays as compared to weekends") +
      scale_x_continuous(breaks = seq(0,2400,100))+
      theme(title = element_text(size = rel(1)))
```

From the graph you can infer the person tends to get up and go to bed earlier during weekdays. Another tendency is that person walks more during the day (10:00-20:00).
Finally, you can see a strong spike in activity in the morning, around 8:35. The
absence of such spike in the evening suggests it may not be generated by 
rushing to workplace, rather by physical activity, let's say jogging.